Efficient Computation of Extended Sources

### 2021-06-07 repro/papers/sseprecond/paper.tex
getting there

${\cal S}^{\pm}$ is not stably invertible (equation \ref{eqn:lamnull}). The
diagonal components of ${\cal S}^{\pm}$ thus carry essentially all of
its information, and it is in terms of these that a sensible inverse problem
is defined.

@@

With these conventions, we can write the version of the source
subproblem studied in this paper as
\begin{equation}
  \label{eqn:esisp}
  \mbox{find }h_s\mbox{ to minimize }\|S^{+}_{z_s,z_r}h_s- d\|^2 +
  \alpha^2\|Ah_s\|^2.
\end{equation}

@@

It follows from the adjoint state method (see Derivations section
below for details) that
\begin{equation}
  \label{eqn:sadj1}
  ({\cal S}^{\pm}_{z_s, z_r})^T = -{\cal S}^{\mp}_{z_r,z_s}
\end{equation}

@@

Since the
mechanical parameters in the homogeneous and lens models are the same
for $z<z_s$, and no rays return to this zone in either model, these
data are asymptotically the same for both models, and I show only the
homogenous medium results. Regard these gathers as the trace
$(P_sp^+,P_sv^+_z)$ on $z=z_s$ of a downgoing acoustic field in $z>z_s$.
Equations \ref{eqn:tracejump10} and
\ref{eqn:tracejump20} show that these differ by a factor of -2 from
source functions $f_s$ and $h_s$ as in the system \ref{eqn:awepm},
with $h_s=0$ and $f_s=0$ respectively. For a point source at $z_d=3500$ m, $x_d=3500$ m, same bandpass filter
wavelet as used for prior examples, causal data on $z=z_s=3000$ m are
shown in Figures \ref{fig:dsrcpghh0} and \ref{fig:dsrcvzghh0}.

@@

Recall that the source vector $(h_s,f_s)$ is assumed to produce a
downgoing field $(p^+,\bv^+)$, that is, emanates high-frequency energy only along
rays that make an angle with the vertical bounded below by a common
minimum angle. Such rays leave $\Omega$ within a common maximum
time. Consequently (Appendix B), in the
slab $z_s<z<z_r$, the field $(p^+,\bv^+)$ approximates the solution of an
anti-causal evolution equation. Choose $\chi(t)$ to be a smooth function
that is $= 0$ for $t \gg 0$ and $=1$ at times when near rays carrying
high-frequency energy in $(p^+,\bv^+)$ cross $z=z_r$. Define 
$(\tilde{p}^-,\tilde{\bv}^-)$ to be the solution in the half-space
$\Omega \times \bR$ of
\begin{eqnarray}
\label{eqn:revawe}
  \frac{1}{\kappa}\frac{\partial \tilde{p}^-}{\partial t} & = & - \nabla \cdot \tilde{\bv}^-, \nonumber \\
  \rho\frac{\partial \tilde{\bv}^-}{\partial t} & = & - \nabla \tilde{p}^-,\nonumber \\
  \tilde{p}^- & =& 0,  \mbox{ for } t \gg 0\\ 
  \tilde{\bv}^- & = & 0 \mbox{ for } t \gg 0\\
  P_r\tilde{p}^- &=& \chi P_rp^+ . 
\end{eqnarray}
That is, $\tilde{p}^-$ has the same boundary value on $z=z_r$ as
$p^+$, except for low-frequency residue that is muted by
$\chi$. Therefore
$p^+ \approx \tilde{p}^-, \bv^+ \approx \tilde{\bv}^-$ near
$z=z_r$. Since the right-hand sides in the system \ref{eqn:awepm} are
singular only on $z=z_s$, and the high-frequency components of
$(p^+,\bv^+)$ are carried by downgoing rays, these differ negligibly
from the the high-frequency components of
$(\tilde{p}^-,\tilde{\bv}^-)$ in the space-time slab $z_s<z<z_r$, and
the approximation holds throughout this region. In particular
$P_sv^+_z \approx P_s \tilde{v}^-_z$. In view of the relation
\ref{eqn:tracejump10},
\begin{equation}
  \label{eqn:tildevtohsubs}
  -2P_s\tilde{v}^-_z \approx h_s,
\end{equation}
so solution
of the system \ref{eqn:revawe} followed by restriction to $z=z_s$ and
multiplication by $-2$ 
approximately inverts the map $S^+{z_s,z_r}: h_s \mapsto P_rp^+$.

Next observe that in view of the relation \ref{eqn:tracejump20}, and
the downgoing nature of the ray system carrying the high frequency
energy in $(p^+,\bv^+)$, the field $(\tilde{p}^-,\tilde{\bv}^-)$ is
actually the restriction to $z<z_r$ of the anti-causal solution of \ref{eqn:awepm}
with $z_s$ replaced by $z_r$, zero constitutive defect, and vertical
load given by the jump in pressure at $z=z_r$ - for this field, use
the same notation. Continuity of vertical
velocity $\tilde{v}^-_z$ at $z=z_r$ implies that the vertical load is
\[
  f_r = -[\tilde{p}^-]|_{z=z_r} = 2 P_r \tilde{p}^- = 2 P_r p^+
\]
(from the definition \ref{eqn:defsamp}, $P_r$ is the limit from the
left). Thus
\[
  P_s \tilde{v^-_z} \approx V^-_{z_r,z_s}(2 P_rp^+) \approx
  2V^-_{z_r,z_s}S^+_{z_r,z_s}h_s.
\]
so
\[
  h_s \approx -2 P_s v^+_z \approx -2 P_s \tilde{v}^-_z \approx
  -4V^-_{z_r,z_s}S^+_{z_r,z_s}h_s
\]
Combine this observation with \ref{eqn:tildevtohsubs} to obtain
\[
 -4  V^-_{z_r,z_s} S^+_{z_s,z_r}  \approx  I,
\]
This relation combines with the identity \ref{eqn:trtrcomp} to
yield the first main result of this section:
\begin{eqnarray}
  \label{eqn:approxinv}
  (V^+_{z_s,z_r})^T S^+_{z_s,z_r} & \approx & \frac{1}{4}I, \nonumber\\
  (S^+_{z_s,z_r})^T V^+_{z_s,z_r} & \approx & \frac{1}{4}I, \nonumber\\
  V^+_{z_s,z_r} (S^+_{z_s,z_r})^T & \approx & \frac{1}{4}I, \nonumber\\
  S^+_{z_s,z_r} (V^+_{z_s,z_r})^T & \approx & \frac{1}{4}I.\\
\end{eqnarray}.
The second equation is simply the transpose of the first, and the
last two follow by by an exactly analogous argument using time
reversal and interchange of the roles of $z_s$ and$z_r$.

@@

I have applied this procedure to the pressure gathers shown in Figures
\ref{fig:drecphh0} and \ref{fig:drecplh0}, generated by a point source
at $z_d=3500$ m, $x_s=3500$ m, propagating in the homogeneous and lens
models respectively. The source time dependence is again a [1, 2.5,
10, 12.5] Hz trapezoidal bandpass filter. As noted earlier, in the
region $z < z_s=3000 m$, these fields are downgoing and can be
regarded as the result of either pressure or velocity source at $z=z_s$: the
pressure source $h_s$ (Figure \ref{fig:dhshh0}) is -2 times the
vertical velocity gather depicted in Figure \ref{fig:dsrcvzhh0}, the
velocity source $f_s$ (Figure \ref{fig:dfshh0}) is -2 times the
pressure gather depicted in Figure \ref{fig:dsrcphh0}.

@@

The remainder of this section is devoted to illustrating two important
features of the approximate inversion developed above. The
first is the validity of the inversion approach given in display
\ref{eqn:approxinv} so long as the fields generated by the
model used in simulation and inversion are downgoing. The models
may be quite far from homogeneous, and even generate triplications,
whereas the inversions remain accurate modulo the dip filtering effect
of aperture. Figure \ref{fig:dinvhsll0} shows the pressure source at $z=z_s$
inverted from the data Figure
\ref{fig:drecpll0}, in which a triplication is obvious. Recall that
this data is generated by propagating a point source at $z_d=3500$ m
in the lens model. The inversion uses the same model. Figure
\ref{fig:dinvhsll0} is an aperture-limited version of the pressure
source at $z=z_s=3000$ m
(Figure \ref{fig:dhsll0}) that generates this data by a dip filter, in
the same way that the sources in Figures \ref{fig:dhsll0} and
\ref{fig:dinvhsll0} are related. The inverted source is an accurate
inversion: it generates the data depicted in Figure
\ref{fig:drerecpll0}, which is very close to the input data (Figure
\ref{fig:drecpll0}). The difference is displayed on the same color
scale in Figure \ref{ddiffrecpll0}.

The second feature is a decoupling of the models used (implicitly or
otherwise) in simulation and those used in inversion: it is possible
to {\em accurately invert pressure data for pressure source using the
  wrong model}, in the sense that the inverted source will
generate an accurate recovery of the input data re-simulated in the
same (wrong) model. For example, inversion of the data of the previous
example (Figure \ref{fig:drecpll0}) {\em in the homogeneous model}
results in an inverted source gather at $z=z_s$ (Figure \ref{fig:dinvhslh0}
that is very far from a dip-filtered version of the source used
generate the data \ref{eqn:dhshh0}. However, this result is an
accurate inversion: re-simulation {\em using the same (homogeneous)
  model as used in the inversion} results in accurate recovery of the
input pressure gather (Figure \ref{fig:drecpll0}). The difference is
shown on the same color scale in Figure \ref{fig:ddiffrecplh0}.

\section{Computing $\Lambda$}

Computation of $\Lambda^{\pm}_{z_s}$ will prove a critical step in an
efficient iterative solution of the linear subproblem
\ref{eqn:esis}. Direct computation of the pressure-to-source operator $\Lambda^{\pm}_{z_s}$, for instance by solving
\ref{eqn:awepm} and reading off $P_sv^{\pm}_{z}$, turns out to be
numerically ill-behaved. The relation \ref{eqn:snull} provides and
alternative approach, taking advantage of the accurate approximate
inverse to $S^+_{z_s,z_r}$ is available. For instance, the first row
of the vector equation \ref{eqn:snull} reads (slightly rearranged):
\[
\Pi_0{\cal S}^+_{z_s,z_r}\Pi_1^T f_s \approx -\frac{1}{2}\Pi_0{\cal
  S}^+_{z_s,z_r}\Pi_0\Lambda^+_{z_s}f_s
\]
\begin{equation}
  \label{eqn:lamidea}
  = -\frac{1}{2} S^+_{z_s,z_r}\Lambda^+_{z_s}f_s.
\end{equation}

The approximate inverse construction for $S^++_{z_s,z_r}$ would permits (approximate)
solution of this equation for $\Lambda^+_{z_s}f_s$: apply
$4(V^{+}_{z_s,z_r})^T$ to both sides of equation \ref{eqn:lamidea} and
use the first equation in the list \ref{eqn:approxinv} to get
\begin{equation}
  \label{eqn:lamident}
  \Lambda^+_{z_s} \approx -8(V^{+}_{z_s,z_r})^T\Pi_0 {\cal S}^+_{z_s,z_r}\Pi_1^T.
\end{equation}
This identity is the second major result of this section: it shows how
to compute that action of $\Lambda^+_{z_s}$ by propagating the input
pressure trace, identified as a source for the velocity evolution,
forward in time from $z_s$ to $z_r$
reading off the pressure trace on $z=z_r$, identifying it once more as
a point load (source for velocity), propagating it backwards in time from $z_r$ to
$z_s$, and finally reading off the velocity trace, interpreted as a
pressure evolution source on $z_s$.

>>>>

${\cal S}^{\pm}$ is not stably invertible: its columns are
approximately linearly dependent, as will be verified below. The
diagonal components of ${\cal S}^{\pm}$ thus carry essentially all of
its information, and it is in terms of these that a sensible inverse problem
is defined.

@@

With these conventions, we can write the version of the source
subproblem \ref{eqn:esis} studied in this paper as
\begin{equation}
  \label{eqn:esisp}
  \mbox{find }h_s\mbox{ to minimize }\|S^{+}_{z_s,z_r}h_s- d\|^2 +
  \alpha^2\|Ah_s\|^2.
\end{equation}

@@

It follows from the adjoint state method (see Appendix A for details) that
\begin{equation}
  \label{eqn:sadj1}
  ({\cal S}^{\pm}_{z_s, z_r})^T = -{\cal S}^{\mp}_{z_r,z_s}
\end{equation}

@@

Since the
mechanical parameters in the homogeneous and lens models are the same
for $z<z_s$, and no rays return to this zone in either model, these
data are asymptotically the same for both models, and I show only the
homogenous medium results. Regard these gathers as the trace
$(P_sp^+,P_sv^+_z)$ on $z=z_s$ of a downgoing acoustic field in $z>z_s$.
Equations \ref{eqn:tracejump10} and
\ref{eqn:tracejump20} show that these differ by a factor of -2 from
source functions $f_s$ and $h_s$ as in the system \ref{eqn:awepm},
with $h_s=0$ and $f_s=0$ respectively. For a point source at $z_d=3500$ m, $x_d=3500$ m, same bandpass filter
wavelet as used for prior examples, causal data on $z=z_s=3000$ m are
shown in Figures \ref{fig:dsrcphh0} and \ref{fig:dsrcvzhh0}.

@@

Recall that the source vector $(h_s,f_s)$ is assumed to produce a
downgoing field $(p^+,\bv^+)$, that is, emanates high-frequency energy only along
rays that make an angle with the vertical bounded below by a common
minimum angle. Such rays leave $\Omega$ within a common maximum
time. Consequently (Appendix B), in the
slab $z_s<z<z_r$, the field $(p^+,\bv^+)$ approximates the solution of an
anti-causal evolution equation. Choose $\chi(t)$ to be a smooth function
that is $= 0$ for $t \gg 0$ and $=1$ at times when near rays carrying
high-frequency energy in $(p^+,\bv^+)$ cross $z=z_r$. Define 
$(\tilde{p}^-,\tilde{\bv}^-)$ to be the solution in the half-space
$\Omega \times \bR$ of
\begin{eqnarray}
\label{eqn:revawe}
  \frac{1}{\kappa}\frac{\partial \tilde{p}^-}{\partial t} & = & - \nabla \cdot \tilde{\bv}^-, \nonumber \\
  \rho\frac{\partial \tilde{\bv}^-}{\partial t} & = & - \nabla \tilde{p}^-,\nonumber \\
  \tilde{p}^- & =& 0,  \mbox{ for } t \gg 0\\ 
  \tilde{\bv}^- & = & 0 \mbox{ for } t \gg 0\\
  P_r\tilde{p}^- &=& \chi P_rp^+ . 
\end{eqnarray}
That is, $\tilde{p}^-$ has the same boundary value on $z=z_r$ as
$p^+$, except for low-frequency residue that is muted by
$\chi$. Therefore
$p^+ \approx \tilde{p}^-, \bv^+ \approx \tilde{\bv}^-$ near
$z=z_r$. Since the right-hand sides in the system \ref{eqn:awepm} are
singular only on $z=z_s$, and the high-frequency components of
$(p^+,\bv^+)$ are carried by downgoing rays, these differ negligibly
from the the high-frequency components of
$(\tilde{p}^-,\tilde{\bv}^-)$ in the space-time slab $z_s<z<z_r$, and
the approximation holds throughout this region. In particular
$P_sv^+_z \approx P_s \tilde{v}^-_z$. In view of the relation
\ref{eqn:tracejump10},
\begin{equation}
  \label{eqn:tildevtohsubs}
  -2P_s\tilde{v}^-_z \approx h_s,
\end{equation}
so solution
of the system \ref{eqn:revawe} followed by restriction to $z=z_s$ and
multiplication by $-2$ 
approximately inverts the map $S^+_{z_s,z_r}: h_s \mapsto P_rp^+$.

Next observe that in view of the relation \ref{eqn:tracejump20}, and
the downgoing nature of the ray system carrying the high frequency
energy in $(p^+,\bv^+)$, the field $(\tilde{p}^-,\tilde{\bv}^-)$ is
actually the restriction to $z<z_r$ of the anti-causal solution of \ref{eqn:awepm}
with $z_s$ replaced by $z_r$, zero constitutive defect, and vertical
load given by the jump in pressure at $z=z_r$ - for this field, use
the same notation. Continuity of vertical
velocity $\tilde{v}^-_z$ at $z=z_r$ implies that the vertical load is
\[
  f_r = -[\tilde{p}^-]|_{z=z_r} =-(\lim_{z\rightarrow
    z_r^+}\tilde{p}^- - \lim_{z\rightarrow
    z_r^-}\tilde{p}^-)
\]
\[
  \approx 2 P_r \tilde{p}^- = 2 P_r p^+
\]
(from the definition \ref{eqn:defsamp}, $P_r$ is the limit from the
left). Thus
\[
  P_s \tilde{v^-_z} \approx V^-_{z_r,z_s}(2 P_rp^+) \approx
  2V^-_{z_r,z_s}S^+_{z_r,z_s}h_s.
\]
so
\[
  h_s \approx -2 P_s v^+_z \approx -2 P_s \tilde{v}^-_z \approx
  -4V^-_{z_r,z_s}S^+_{z_r,z_s}h_s
\]
Combine this observation with \ref{eqn:tildevtohsubs} to obtain
\[
 -4  V^-_{z_r,z_s} S^+_{z_s,z_r}  \approx  I,
\]
This relation combines with the identity \ref{eqn:trtrcomp} to
yield the first main result of this section:
\begin{eqnarray}
  \label{eqn:approxinv}
  (V^+_{z_s,z_r})^T S^+_{z_s,z_r} & \approx & \frac{1}{4}I, \nonumber\\
  (S^+_{z_s,z_r})^T V^+_{z_s,z_r} & \approx & \frac{1}{4}I, \nonumber\\
  V^+_{z_s,z_r} (S^+_{z_s,z_r})^T & \approx & \frac{1}{4}I, \nonumber\\
  S^+_{z_s,z_r} (V^+_{z_s,z_r})^T & \approx & \frac{1}{4}I.\\
\end{eqnarray}.
The second equation is simply the transpose of the first, and the
last two follow by by an exactly analogous argument using time
reversal and interchange of the roles of $z_s$ and$z_r$.

@@

I have applied this procedure to the pressure gathers shown in Figures
\ref{fig:drecphh0} and \ref{fig:drecplh0}, generated by a point source
at $z_d=3500$ m, $x_s=3500$ m, propagating in the homogeneous and lens
models respectively. The source time dependence is again a [1, 2.5,
10, 12.5] Hz trapezoidal bandpass filter. As noted earlier, in the
region $z < z_s=3000 m$, these fields are downgoing and can be
regarded as the result of either pressure or velocity source at $z=z_s$: the
pressure source gather $h_s$ (Figure \ref{fig:dhshh0})  is -2 times the
vertical velocity gather depicted in Figure \ref{fig:dsrcvzhh0}, the
velocity source gather $f_s$ (Figure \ref{fig:dfshh0}) is -2 times the
pressure gather depicted in Figure \ref{fig:dsrcphh0}.

@@

The remainder of this section is devoted to illustrating two important
features of the approximate inversion developed above. The
first is the validity of the inversion approach given in display
\ref{eqn:approxinv} so long as the fields generated by the
model used in simulation and inversion are downgoing. The models
may be quite far from homogeneous, and even generate triplications,
whereas the inversions remain accurate modulo the dip filtering effect
of aperture. Figure \ref{fig:dinvhsll0} shows the pressure source at $z=z_s$
inverted from the data Figure
\ref{fig:drecplh0}, in which a triplication is obvious. Recall that
this data is generated by propagating a point source at $z_d=3500$ m
in the lens model. The inversion uses the same model. Figure
\ref{fig:dinvhsll0} is an aperture-limited version of the pressure
source at $z=z_s=3000$ m
(Figure \ref{fig:dhshh0}) that generates this data by a dip filter, in
the same way that the sources in Figures \ref{fig:dhshh0} and
\ref{fig:dinvhshh0} are related. The inverted source is an accurate
inversion: it generates the data depicted in Figure
\ref{fig:drerecpll0}, which is very close to the input data (Figure
\ref{fig:drecplh0}). The difference is displayed on the same color
scale in Figure \ref{fig:ddiffrecpll0}.

The second feature is a decoupling of the models used (implicitly or
otherwise) in simulation and those used in inversion: it is possible
to {\em accurately invert pressure data for pressure source using the
  wrong model}, in the sense that the inverted source will
generate an accurate recovery of the input data re-simulated in the
same (wrong) model. For example, inversion of the data of the previous
example (Figure \ref{fig:drecplh0}) {\em in the homogeneous model}
results in an inverted source gather at $z=z_s$ (Figure \ref{fig:dinvhslh0}
that is very far from a dip-filtered version of the source used
generate the data \ref{fig:dhshh0}. However, this result is an
accurate inversion: re-simulation {\em using the same (homogeneous)
  model as used in the inversion} results in accurate recovery of the
input pressure gather (Figure \ref{fig:drecplh0}). The difference is
shown on the same color scale in Figure \ref{fig:ddiffrecplh0}.

The next chapter in this story recognizes the relations in display
\ref{eqn:approxinv} as asserting the approximate unitarity of
$S^+_{z_s,z_r}$.

The matrix identity \ref{eqn:snull} implies a relation between $S, V,$
and $\Lambda$ of some interest in itself. After minor re-arrangement, the second row of reads
\begin{equation}
  \label{eqn:snull2}
-\frac{1}{2}\Pi_1{\cal S}^{\pm}_{z_s,z_r}\Pi_0^T\Lambda^{\pm}_{z_s}  \approx
V^{\pm}_{z_s,z_r}.
\end{equation}
In these relations, the projection on the left picks out the vertical velocity component
of a downgoing wavefield at $z=z_r$: that is,
\[
-\frac{1}{2}\Pi_1{\cal S}^{\pm}_{z_s,z_r}\Pi_0^T\Lambda^{\pm}_{z_s}P_sp^+
=-\frac{1}{2}P_r v_z^+,
\]
where $(p^+, \bv^+)$ solve the system \ref{eqn:awepm} with $f_s=0$ and
$h_s = \Lambda^{\pm}_{z_s}P_sp^+$. On the other hand, from relation
\ref{eqn:tracejump10},
\[
  P_r v_z^+ = -\frac{1}{2}\Lambda^+_{z_r}P_r p^+
\]
where
\[
  P_r p^+ = \Pi_0{\cal S}^{+}_{z_s,z_r}\Pi_0^T\Lambda^{+}_{z_s}P_s
  p^+
\]
\[
  = S^+_{z_s,z_r}\Lambda^{+}_{z_s}P_sp^+
\]
Therefore combining the last two equations with \ref{eqn:snull2},
obtain
\begin{equation}
  \label{eqn:sv}
  \frac{1}{4}\Lambda^+_{z_r}S^+_{z_s,z_r}\Lambda^{+}_{z_s} = V^+_{z_s,z_r}.
\end{equation}
This is the promised relation.

A minor extension of an earlier example, based on the gathers
displayed in Figures \ref{fig:dsrcphh0} ($P_sp^+$) and
\ref{fig:dsrcvzhh0} ($P_sv_z^+$) illustrates the identity
\ref{eqn:sv}.
The pressure gathers at $z=z_r$ produced by application of
$S^+_{z_r,z_s}$ to $h_s=-2P_s v^+_z$, and $\Pi_0{\cal S}^+_{z_s,z_r}\Pi_1^T$ to $f_s
=-2P_sp^+$, have already been displayed (Figures \ref{fig:dfwdphh0}
and \ref{fig:daltphh0}), and are identical to several digits. The same 
is true of the vertical velocity gathers, shown in \ref{fig:dfwdvzhh0}
and \ref{fig:daltvzhh0}; their difference, plotted on the same scale,
appears in Figure \ref{fig:dsvcomphh0}. Once again, this relation
expresses the interchangeability of pressure and velocity sources,
related by $\Lambda$.

Relation \ref{eqn:tracejump10} characterizes
$\Lambda^+_{z_s}$ as connecting the pressure and velocity components
of downgoing fields restricted to $z=z_s$. That is, the pressure
source gather
$h_s=\Lambda^{+}_{z_s}P_s p^+ = -2
P_{z_s}v_z$, displayed in Figure \ref{fig:dhshh0}, is the image of the
pressure gather in Figure \ref{fig:dsrcphh0} under
$\Lambda^{+}_{z_s}$.
The pressure gather
\ref{fig:dfwdphh0} is the image of this pressure source gather under
$S^+_{z_s,z_r}$. The corresponding vertical velocity gather
(Figure \ref{fig:dfwdvzhh0}) is $-1/2$ times $\Lambda^+_{z_r}P_r
p$. Therefore scaling the data in Figure \ref{fig:dfwdvzhh0} by
$-2$ produces
$Lambda^+_{z_r}S^+_{z_s,z_r}\Lambda^{+}_{z_s}P_s p^+$. On the other
hand, figure \ref{fig:daltvzhh0} shows the result of applying
$V^+_{z_s,z_r}$ to $f_s=-2P_s p^+$. Therefore scaling the gather in
Figure \ref{fig:daltvzhh0} by $-\frac{1}{2}$ produces $V^+_{z_s,z_r}P_zp^+$.
Since the data in Figures \ref{fig:dfwdvzhh0} and \ref{fig:daltvzhh0}
are essentially identical, the relation \ref{eqn:sv} holds for this
example.

As shown in the last section, $4(V_{z_s,z_r}^+)^T$ is approximately
inverse to $S^{+}_{z_s,z_r}$. Therefore, transposing both sides of
equation \ref{eqn:sv} and using \ref{eqn:approxinv}, obtain
\begin{equation}
  \label{eqn:almostunitary}
  4(V_{z_s,z_r}^+)^TS^+_{z_s,z_r} = [ (\Lambda^+_{z_r})^T
  (S^{+}_{z_s,z_r})^T(\Lambda^+_{z_s})^T]S^{+}_{z_s,z_r} \approx I.
\end{equation}

The remarkable feature of the identity \ref{eqn:almostunitary} is that
it exhibits an approximate right inverse of $S^+$ as an adjoint with
respect to a weighted inner product - or it would, if the operators
$(\Lambda^+)$ were symmetric positive definite. As noted earlier,
these operators are only approximately symmetric, though they are
positive semi-definite. That is not a great obstacle, however:
symmetrizing them in the obvious way commits a negligible error, of
the sort that this paper already neglects wholesale. That is,
\begin{equation}
  \label{eqn:unitary}
  [ \frac{1}{2}((\Lambda^+_{z_r})^T+ \Lambda^+_{z_r})
  (S^{+}_{z_s,z_r})^T \frac{1}{2}((\Lambda^+_{z_s})^T+
  \Lambda^+_{z_s})]S^{+}_{z_s,z_r} \approx I.
\end{equation}
The symmetrized $\Lambda$ operators are at least positive
semi-definite, hence define (at least) semi-norms.

Similar relations have been derived for other scattering operators,
and have been used to accelerate iterative solutions of inverse
scatering problems: \cite{DafniSymes:SEG18b} review some of this
literature.

For convenience, in this section write $S$ in place of
$S^+_{z_s,z_r}$. Also abbreviate the symmetrized $\Lambda$ operators
using notation suggesting weight
operators in model and data spaces:
\begin{eqnarray}
  W_m^{-1}&=& \frac{1}{2}((\Lambda^+_{z_s})^T+
              \Lambda^+_{z_s}),\nonumber \\
  W_d &=& \frac{1}{2}((\Lambda^+_{z_r})^T+ \Lambda^+_{z_r}).
          \label{eqn:wdef}
\end{eqnarray}
The identification of the symmetrized $\Lambda^+_{z_s}$ as the inverse
of another operator $W_m$ is formal, since the former operator is
likely to have null (or nearly-null) vectors due to aperture-related
amplitude loss. I will proceed as if this were sensible, since the
mathematical issues can be defused and the only the application of
$W_m^{-1}$ is required in the algorithms described here.

Adopting Hilbert norms defined by the operators $W_m$ and $W_d$ in its
domain and range respectively, the adjoint of $S$ is given by
\begin{equation}
\label{eqn:wadj}
S^{\dagger} = W_m^{-1}S^TW_d,
\end{equation}

In this notation, the relation \ref{eqn:unitary} takes the form
\begin{equation}
  \label{eqn:wunitary}
  S^{\dagger}S \approx I.
\end{equation}
That is to say, $S$ is approximately unitary with respect to the
weighted norms defined by $W_m$ and $W_d$. Therefore a Krylov space
method employing these norms will converge rapidly, at least for the
well-determined components of the solution.

Similar relations have been derived for other scattering operators,  
and have been used to accelerate iterative solutions of inverse  
scattering problems: \cite{DafniSymes:SEG18b} review some of this  
literature.

The most convenient arrangement the Conjugate Gradient (CG) algorithm
taking advantage of the structure \ref{eqn:wadj} is the {\em
  Preconditioned CG}.

Allowing that the fit error will be measured by the data space norm,
the least squares problem to be solved is not just $Sh \approx d$, but
a regularized version:
\begin{equation}
  \label{eqn:einv}
  \mbox{minimize}_h \|Sh-d\|^2_d + \alpha^2 \|Ah\|^2_m
\end{equation}
Remark: recall that the modified data space norm $\|d\|_d^2 = \langle
d, W_d d\rangle$ has physical meaning: for acoustics, it is
proportional to the power transmitted to the fluid by the source.

The minimizer of the objective defined in equation \ref{eqn:einv}
solves the normal equation
\begin{equation}
  \label{eqn:norm0}
  (S^{\dagger}S + \alpha^2 A^{\dagger}A)h = S^{\dagger}d 
\end{equation}
where the weighted adjoint $S^{\dagger}$ has already been defined in equation \ref{eqn:wadj}, and $A^{\dagger}$ is the adjoint of $A$ in the weighted model space norm defined by $W_m$, namely
\begin{equation}
  \label{eqn:aadj}
  A^{\dagger} = W_m^{-1}A^TW_m.
\end{equation}

Note that the normal operator appearing on the left-hand side of
\ref{eqn:norm0} is not an approximate identity, due to the presence of
the regularization term: the spectrum increases in spread with
increasing $\alpha$, leading to slower convergence. Fortunately for
the present setting, the operators $W_m^{-1}$, $A$, and $W_m$
approximately commute (they are scalar {\em pseudodifferential}, once
the difficulties with the definition of $W_m$, mentioned above, are
taken care of), so that $A^{\dagger} \approx A^T$. Therefore
\begin{equation}
  \label{eqn:normapprox}
  S^{\dagger}S + \alpha^2 A^{\dagger}A \approx I + \alpha^2A^TA
\end{equation}
Recall that $A$ is simply multiplication by the Euclidean distance to
the physical source point $\bx_s$: $A u (\bx) = |\bx-\bx_s|u(\bx),
A^TAu(\bx) = |\bx-\bx_s|^2u(\bx)$. So the equation $(I+\alpha^2
A^TA)u=b$ is trivial to solve, and this is a key characteristic of a
good preconditioner. However this observation must be combined with
the wegithed norm structure.

Rewrite the normal equation \ref{eqn:norm} as
\begin{equation}
  \label{eqn:norm1}
  W_m^{-1}(S^TW_dS + \alpha^2 A^TW_mA)h = W_m^{-1}S^TW_md 
\end{equation}
Since $W_m$ is self-adjoint and positive definite, the common factor on both sides of \ref{eqn:norm1} can be re-written as
\begin{equation}
  \label{eqn:normpart}
  Nh = (S^*S + \alpha^2 A^*A)h = S^*d 
\end{equation}
in which $S^*, A^*$ are the adjoints with the original (Euclidean)
inner product in the domains but the weighted inner product in data
space:
\begin{eqnarray}
  \label{eqn:sadj}
  S^* &=& S^T W_d,\\
  A^* &=& A^T W_m.
\end{eqnarray}
Note the $S^*S$ and $A^*A$ are symmetric in the Euclidean sense, so
equation \ref{eqn:normpart} is a symmetric positive (semi-)definite
linear system, just the sort of thing for which the 
The Preconditioned Conjugate Gradient (``PCG'') algorithm was
designed. PCG for solution
of equation \ref{eqn:normpart} with preconditioner $M$is usually
written as Algorithm 1 (see for example \cite{Golub:2012}):

\begin{algorithm}[H]
\caption{Preconditioned Conjugate Gradient Algorithm, Standard Version}
\begin{algorithmic}[1]
\State Choose $h_0=0$ 
  \State $r_0 \gets S^*d$
  \State $p_0 \gets M^{-1} r_0$
  \State $g_0 \gets p_0$
  \State $q_0 \gets Np_0$
  \State $k \gets 0$
  \Repeat
  \State $\alpha_k \gets \frac{\langle g_k,r_k \rangle}{\langle p_k,q_k\rangle}$
  \State $h_{k+1} \gets h_k + \alpha_k p_k$
  \State $r_{k+1} \gets r_k - \alpha_kq_k$
  \State $g_{k+1} \gets M^{-1} r_{k+1}$
  \State $\beta_{k+1} \gets \frac{\langle g_{k+1},r_{k+1}\rangle}{\langle g_k,r_k\rangle}$
  \State $p_{k+1}\gets g_{k+1}+\beta_{k+1}p_k$
  \State $q_{k+1} \gets Np_{k+1}$
  \State $k \gets k+1$
  \Until{Error is sufficiently small, or max iteration count exceeded} 
\end{algorithmic}
\end{algorithm}
The iteration converges rapidly if $M^{-1}N \approx I$. This is true
if and only if the symmetrized operator $M^{-1/2}NM^{-1/2} \approx I$,
which is in turn true if the eigenvalues of $M^{-1/2}NM^{-1/2}$ are
close to 1 (actually works well is most of these eigenvalues are close
to 1, and the rest are small - which is the case for the current
problem)..  Further, PCG is computationally effective is M is easy to
invert.

From \ref{eqn:normapprox} and \ref{eqn:norm1}, it follows that
\[
  W_m^{-1}(S^TW_dS + \alpha^2 A^TW_mA) \approx I + \alpha^2 A^TA.
\]
This observation suggests using $M=W_m(I+\alpha^2A^TA)$. This choice
is not symmetric, but since the operators on the right-hand side are
scalar pseudodifferential hence commute, it is equivalent to use of
\begin{eqnarray}
  M         &=&(I+\alpha^2A^TA)^{1/2}W_m(I+\alpha^2A^TA)^{1/2},\nonumber \\
  M^{-1}
            &=&(I+\alpha^2A^TA)^{-1/2}W_m^{-1}(I+\alpha^2A^TA)^{-1/2}.
                \label{eqn:defprecond}
\end{eqnarray}
 With this choice, \ref{eqn:normapprox} implies that
$M^{-1}N \approx I$, also $M$ is symmetric. As already mentioned,
powers of $I + \alpha^2A^TA$ are trivial to compute, given the choice
of $A$ made here. We will examine fast algorithms for computing
$W_m^{-1}$ = the symmetrized pressure-to-source operator in the next
section. Note that only $M^{-1}$, hence only $W_m^{-1}$, appears in
Algorithm 1.

[Example]

\section{Computing and Symmetrizing $\Lambda$}

Computation of $\Lambda^{\pm}_{z_s}$ will prove a critical step in an
efficient iterative solution of the linear subproblem
\ref{eqn:esis}. Direct computation of the pressure-to-source operator $\Lambda^{\pm}_{z_s}$, for instance by solving
\ref{eqn:awepm} and reading off $P_sv^{\pm}_{z}$, turns out to be
numerically ill-behaved. The relation \ref{eqn:snull} provides and
alternative approach, taking advantage of the accurate approximate
inverse to $S^+_{z_s,z_r}$ is available. The first row of
\ref{eqn:snull} is given explicitly in \ref{eqn:snull1}, which I
repeat for the reader's convenience:
\begin{equation}
  \label{eqn:lamidea}
  \Pi_0{\cal S}^+_{z_s,z_r}\Pi_1^T f_s \approx  -\frac{1}{2} S^+_{z_s,z_r}\Lambda^+_{z_s}f_s.
\end{equation}
